---
title: ""
author: "Justin Chow"
date: "3/10/2021"
output: html_document
---

```{r setup, echo=FALSE, message=F, warning=F}
# Clear the working space
rm(list = ls())

# Set working directory
setwd("C:/Users/14087/Documents/github/airline-covid19")

library(AER)
library(doBy)
library(dplyr)
library(ggplot2)
library(psych)
library(readxl)
library(stargazer)
library(tidyr)
library(zoo)

```

# Data

Our data source contains data on 14 U.S. and foreign airline companies as well as several airports within the United States. The dataset contains information on several variables including passenger count, flights, net income, operating revenue and more during the time period of October 2002 to November 2020. There is data on each of these variables separated by month as well as domestic or international flights. We downloaded data for Alaska Airlines on 10 major airports.

```{r data, echo=FALSE, warning=FALSE, message=FALSE}

# Load 10 randomly assigned Major Airports for Alaska Airlines
den <- readxl::read_excel("DEN.xls.xlsx", col_names = c("Year", "Month", "Domestic", "International", "Total"), skip = 2)
bwi <- readxl::read_excel("BWI.xls.xlsx", col_names = c("Year", "Month", "Domestic", "International", "Total"), skip = 2)
dtw <- readxl::read_excel("DTW.xls.xlsx", col_names = c("Year", "Month", "Domestic", "International", "Total"), skip = 2)
slc <- readxl::read_excel("SLC.xls.xlsx", col_names = c("Year", "Month", "Domestic", "International", "Total"), skip = 2)
pdx <- readxl::read_excel("PDX.xls.xlsx", col_names = c("Year", "Month", "Domestic", "International", "Total"), skip = 2)
phl <- readxl::read_excel("PHL.xls.xlsx", col_names = c("Year", "Month", "Domestic", "International", "Total"), skip = 2)
mco <- readxl::read_excel("MCO.xls.xlsx", col_names = c("Year", "Month", "Domestic", "International", "Total"), skip = 2)
msp <- readxl::read_excel("MSP.xls.xlsx", col_names = c("Year", "Month", "Domestic", "International", "Total"), skip = 2)
iah <- readxl::read_excel("IAH.xls.xlsx", col_names = c("Year", "Month", "Domestic", "International", "Total"), skip = 2)
hnl <- readxl::read_excel("HNL.xls.xlsx", col_names = c("Year", "Month", "Domestic", "International", "Total"), skip = 2)

# Rename assigned Major Airports data sets
den$origin <- "DEN"
bwi$origin <- "BWI"
dtw$origin <- "DTW"
slc$origin <- "SLC"
pdx$origin <- "PDX"
phl$origin <- "PHL"
mco$origin <- "MCO"
msp$origin <- "MSP"
iah$origin <- "IAH"
hnl$origin <- "HNL"

# Total number of Flights and Passengers for Alaska Airlines
flights <- readxl::read_excel("FLIGHTS.xls.xlsx", col_names = c("Year", "Month", "Domestic", "International", "Total"), skip = 2)
passengers <- readxl::read_excel("PASSENGERS.xls.xlsx", col_names = c("Year", "Month", "Domestic", "International", "Total"), skip = 2)

# Bind 10 randomly assigned airports into one variable
major <- rbind(bwi,den,dtw,hnl,iah,mco,msp,pdx,phl,slc)

```

# Analysis

We randomly selected 10 major airports that Alaska Airlines flies to within the United States. These airports include:

- Portland
- Salt Lake City
- Honolulu
- Denver
- Houston
- Orlando
- Minneapolis
- Detroit
- Baltimore
- Philadelphia

Alaska Airlines is headquartered in Seattle, Washington so there is considerably more travel to and from Portland than any other airport due to its proximity to Seattle.

We omitted Portland from the majority of our analysis as it significantly skewed the data. Alaska already has a prominent presence at Phoenix, so we decided to focus on flight demand for our other chosen airports.

```{r major, echo=FALSE, warning=FALSE, message=FALSE}

# Covert the variables in the major data set everything to lower case
names(major) <- tolower(names(major))
major <- major[major$month != "TOTAL", ]
major <- major[order(major$year,decreasing=FALSE),]

# Format year and month to create datetime variable 
major$date <- as.yearmon(paste(major$year, major$month), "%Y %m")
major <- major[order(major$date,decreasing=FALSE),]

```


```{r table, echo=FALSE, message=F, warning=F, comment=" "}

# Return summary statistics and describe the data frame
summary(major)
describe(major)

```


```{r image1, echo=FALSE, message=F, warning=F}

# Time Series plot in domestic flights for all flights from all major airports
ggplot(major, aes(x=date, y=domestic, colour = origin)) + geom_line() +
  labs(title = "All Major Airline Flights", x = "Dates", y = "Flights") 

```


```{r image2, echo=FALSE, message=F, warning=F}

# Subset major airlines to omit the outlier flights from Phoenix airport
major_all <- subset(major, origin != 'PDX')

# Time Series plot in domestic flights for all flights from major airports except from Phoenix
ggplot(major_all, aes(x=date, y=domestic, colour = origin)) + geom_line() +
  labs(title = "All Major Airline Flights", x = "Dates", y = "Flights") 

```


```{r image2, echo=FALSE, message=F, warning=F}

# Subset flights from major airlines that occur from the beginning of 2019 to the end of 2020
subset1 <- subset(major_all, year >= 2019 & year <= 2020)

# Stacked bar chart in domestic flights for all flights from major airports except from Phoenix
ggplot(subset1, aes(x=date)) + 
  geom_bar(aes(y = domestic/sum(domestic), fill=origin), stat="identity") +
  labs(title = "All Major Flights 2019-2020", x = "Dates", y = "Percent Change") 

```
